R, Python, and Ruby clients for GBIF species occurrence data

### 2020-02-27 manuscript_mee.Rmd
tweaks to mee paper

---
layout: review, 11pt
linenumbers: true
title: "R, Python, and Ruby clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      rOpenSci, Department of Enivornmental Science, Policy and Management, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

      Email address: [scott@ropensci.org](mailto:scott@ropensci.org)

@@

      Discussion. Programmatic access to GBIF will facilitate more open
        and reproducible science - the three GBIF clients described herein
        are a significant contribution towards this goal.

bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle

output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

The `rgbif` package is nearly completely written in R (a small Javascript
library is included for reading well known text [@wkt]), uses an [MIT license][mit]
to maximize use everywhere. `rgbif` is developed publicly on GitHub
at [https://github.com/ropensci/rgbif][ghrgbif], where development versions
of the package can be installed, and bugs and feature requests reported.
Stable versions of `rgbif` can be installed from [CRAN][cranrgbif], the
distribution network for R packages. `rgbif` is part of the rOpenSci project
(http://ropensci.org), a developer network making R software to facilitate
reproducible research.

@@

### Library interfaces

@@

### Datasets

@@

#### Dataset metrics

@@

### Networks, nodes, and installations

@@

### GBIF Backbone

@@

### Name searching

@@

### Download API

@@

#### Downloaded data format

@@

### Search API

@@

#### Search for occurrences

@@

#### Data cleaning

@@

## Mapping

An obvious downstream use case for species occurrence data is to map the
data. `rgbif` per se is largely not concerned with making this easier,
although we do have a simple wrapper around `ggplot2` to make it easy to
get a quick plot of occurrence data. For example, here we plot 100
occurrences for _Puma concolor_.

```{r fig.width=8, fig.height=4}
key <- name_backbone(name='Puma concolor')$speciesKey
dat <- occ_search(taxonKey = key, limit = 100, hasCoordinate = TRUE)
gbifmap(dat$data)
```

Another package, [mapr][mapr], is the perfect mapping companion to `rgbif`.
It has convenient functions for handling input data from `rgbif`, `spocc`,
or arbitrary data.frame's, and output plots for base plots, `ggplot2`,
`ggmap` (`ggplot2` with map layers underneath), and interactive maps on
GitHub gists or with Leaflet.js.

@@

## GBIF data in other R packages

@@

### taxize

@@

### spocc

@@

### Ecological niche modeling

@@

### Biodiversity in big cities

@@

### Valley oak occurrence data comparison

@@

The `rgbif`, `pygbif`, and `gbibrb` libraries provide programmatic interfaces
to GBIF's application programming interface (API) - a powerful tool for
working with species occurrence data, and facilitating reproducible research.
In fact, the `rgbif` package has already been used in more than 20 scholarly
publications (as of 2016-08-10), including [@Amano_2016, @Bartomeus_2013, @Barve_2014, @Bone_2015, @Collins_2015, @Drozd_2013, @Kong_2015, @Richardson_2015, @Turner_2015, @Verheijen_2015, @Zizka_2015, @Butterfield_2016, @Dellinger_2015, @Feitosa_2015, @Malhado_2015, @Werner_2015, @Robertson_2016, @Davison_2015, @Janssens_2016].

The `rgbif` package is relatively stable, and should not have
many breaking changes unless necessitated due to changes in the
GBIF API. However, it will gain function(s) to work with the maps
API in the near future.

The `pygbif` and `gbifrb` libraries are in early development, and will
greatly benefit from any feedback and use cases.

@@

[mit]: http://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: http://www.gbif.org/developer/summary
[gbifissues]: http://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html
[taxize]: https://github.com/ropensci/taxize
[spocc]: https://github.com/ropensci/spocc
[mapr]: https://github.com/ropensci/mapr
[ghpygbif]: https://github.com/sckott/pygbif
[pypipygbif]: https://pypi.python.org/pypi/pygbif

>>>>

---
preamble: |
  \journal{Methods in Ecology and Evolution}
  \linenumbers
  \usepackage{setspace}
  \doublespacing
bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
output:
  rticles::elsevier_article:
    keep_tex: true
    latex_engine: xelatex
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
title: "R, Python, and Ruby clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: cboettig(at)berkeley.edu
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      rOpenSci, Department of Enivornmental Science, Policy and Management, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

      Email address: [myrmecocystus@gmail.com](mailto:myrmecocystus@gmail.com)

@@

      Discussion. Programmatic access to GBIF will facilitate more open
        and reproducible science - the three GBIF clients described herein
        are a significant contribution towards this goal.
---

@@

The `rgbif` package is nearly completely written in R (a small Javascript
library is included for reading well known text [@wkt]), uses an [MIT license][mit]
to maximize use everywhere. `rgbif` is developed publicly on GitHub
at [https://github.com/ropensci/rgbif][ghrgbif], where development versions
of the package can be installed, and bugs and feature requests reported.
Stable versions of `rgbif` can be installed from [CRAN][cranrgbif], the
distribution network for R packages. `rgbif` is part of the rOpenSci project
(https://ropensci.org), a developer network making R software to facilitate
reproducible research.

@@

# Library interfaces

@@

## Datasets

@@

## Dataset metrics

@@

## Networks, nodes, and installations

@@

## GBIF Backbone

@@

## Name searching

@@

## Occurrences Download API

@@

### Downloaded data format

@@

## Search API

@@

### Search for occurrences

@@

# Data cleaning

@@

# Mapping

An obvious downstream use case for species occurrence data is to map the
data. `rgbif` has enough to do without handling mapping. Another package, [mapr][],
works with `rgbif` to make this easy. For example, here we plot 100
occurrences for _Puma concolor_.

```{r fig.width=8, fig.height=4}
library(mapr)
key <- name_backbone(name='Puma concolor')$speciesKey
dat <- occ_search(taxonKey = key, limit = 100, hasCoordinate = TRUE)
map_ggplot(dat, name = "species")
```

`mapr` has convenient functions for handling input data from `rgbif`, `spocc`,
or arbitrary data.frame's, and output plots for base plots, `ggplot2`,
`ggmap` (`ggplot2` with map layers underneath), and interactive maps on
GitHub gists or with Leaflet.js.

@@

# Citing GBIF data

All the data within GBIF was painstakingly collected by people and institutions
across the globe. They all, including GBIF, deserve to be cited appropropriately
if you use any GBIF data in your research. There are clear forces making this
difficult however. Publishers often limit, without reason if the publication is
online only, the number of citations one can include a manuscript. Even when 
this happens, you can always include additional citations in an appendix.

To make citing data from GBIF easier `rgbif` has a function `gbif_citation()`,
which accepts the output of many different `rgbif` functions and gives
citations for each dataset represented in the data therein. For example,
suppose you searched for occurrences for `taxonKey` 9206251 (_Helianthus annuus_):

```{r}
res <- occ_search(taxonKey = 9206251, limit = 20)
```

We retrieved `r NROW(res$data)` occurrences from `r length(unique(res$data$datasetKey))`
different datasets.

To get citations for each of the four datasets, one can pass the output of `occ_search()`,
here the object `res`, to the `gbif_citation()` function:

```{r}
cites <- gbif_citation(res)
cites
```

And we get four citations, one of each of the datasets behind the `r NROW(res$data)`
occurrence records retrieved.

We don't yet have the same functionality available in `pygbif`, but it will be
coming soon.

## Downloads vs. searches

The distinction between "downloads" and "searches" is not especially meaningful with
respect to the science done with GBIF data. However, downloads and searches
are different with respect to easily being able to cite GBIF data, which 
matters for GBIF and for data providers.

Downloads refers to the `rgbif` functions that start with `occ_download`, and the
`pygbif` methods that start with `occurrences.download`. These methods use the GBIF
"downloads API", which you can interact with from `rgbif` and `pygbif` programatically,
similar to how you would interact with the GBIF website - each interaction creating
a download request, which processes in the background. The user has to ask if 
the request is ready to download. Downloads can request a very large amount of data.

Searches are faster and more dynamic. Search functions in `rgbif` include `occ_search`,
`occ_data`, and `occ_get`, while `pygbif ` methods include `occurrences.search`, 
`occurrences.get`. These methods can only request up to 200,000 occurrence records.

The big difference between downloads and searches with respect to citations is that
each download generates a Digital Object Identifier (or DOI). This DOI can be used
cite the entire set of data, that can inlude many datasets. There's no good way to
refer to the entire results of searches, other than the user manually creating a 
dataset somewhere public and linking to it from Zenodo to get a DOI. However,
simply using downloads is simpler in terms of getting a single citeable DOI for
the data used in a research paper/project.

In `rgbif`, the output of `occ_download_meta()` can be passed to `gbif_citation()`
to get the citation for the entire download.

```{r eval=FALSE}
key <- "0000122-171020152545675"
res <- occ_download_meta(key)
gbif_citation(res)
#> $download
#> [1] "GBIF Occurrence Download https://doi.org/10.15468/dl.yghxj7 Accessed from R via rgbif (https://github.com/ropensci/rgbif) on 2017-10-20"
#> 
#> $datasets
#> NULL
```

If you however download the data first, then `gbif_citation()` can get citations 
for all the datasets contained within the download.

```{r eval=FALSE}
d1 <- occ_download_get(key, overwrite = TRUE)
gbif_citation(d1)
#> $download
#> [1] "GBIF Occurrence Download https://doi.org/10.15468/dl.yghxj7 Accessed from R via rgbif (https://github.com/ropensci/rgbif) on 2017-10-20"
#> 
#> $datasets
#> $datasets[[1]]
#> <<rgbif citation>>
#>    Citation: Grant S, Jones J (2017). Field Museum of Natural History (Zoology)
#>         Invertebrate Collection. Version 18.6. Field Museum. Occurrence Dataset
#>         https://doi.org/10.15468/6q5vuc accessed via GBIF.org on 2017-10-20..
#>         Accessed from R via rgbif (https://github.com/ropensci/rgbif) on
#>         2020-02-24
#>    Rights: To the extent possible under law, the publisher has waived all
#>         rights to these data and has dedicated them to the Public Domain (CC0
#>         1.0). Users may copy, modify, distribute and use the work, including
#>         for commercial purposes, without restriction.
#> 
#> $datasets[[2]]
#> <<rgbif citation>>
#>    Citation: Creuwels J (2017). Naturalis Biodiversity Center (NL) - Crustacea.
#>         Naturalis Biodiversity Center. Occurrence Dataset
#>         https://doi.org/10.15468/vjoltu accessed via GBIF.org on 2017-10-20..
#>         Accessed from R via rgbif (https://github.com/ropensci/rgbif) on
#>         2020-02-24
#>    Rights: To the extent possible under law, the publisher has waived all
#>         rights to these data and has dedicated them to the Public Domain (CC0
#>         1.0). Users may copy, modify, distribute and use the work, including
#>         for commercial purposes, without restriction.
```

# GBIF data in other R packages

@@

## taxize

@@

## spocc

@@

## Ecological niche modeling

@@

## Biodiversity in big cities

@@

## Valley oak occurrence data comparison

@@

The `rgbif`, `pygbif`, and `gbibrb` libraries provide programmatic interfaces
to GBIF's application programming interface (API) - a powerful tool for
working with species occurrence data, and facilitating reproducible research.
In fact, the `rgbif` package has already been used in more than 20 scholarly
publications (as of 2020-02-25), for example @Amano_2016, @Bartomeus_2013,
and @Werner_2015 (see Appendix A for all citations).

The `rgbif` package is stable, and should not have
many breaking changes unless necessitated due to changes in the
GBIF API. The `pygbif` and `gbifrb` libraries are in early development,
and will greatly benefit from any feedback and use cases.

@@

[mit]: https://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: https://www.gbif.org/developer/summary
[gbifissues]: https://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html
[taxize]: https://github.com/ropensci/taxize
[spocc]: https://github.com/ropensci/spocc
[mapr]: https://github.com/ropensci/mapr
[ghpygbif]: https://github.com/sckott/pygbif
[pypipygbif]: https://pypi.python.org/pypi/pygbif
[gemgbif]: https://rubygems.org/gems/gbifrb
